The Post-2007 Declining U.S. Birth Rates imply Significant Societal Shifts in the U.S.

### 2023-02-22 outputs/paper/paper.qmd
Add files via upload

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-2
#| fig-cap: Race and Ethnicity (ages 15-19 from 2000 to 2020)

@@

@@

Figure 5. place it here with the Figure 2-a from the Original paper (@fig-5).

@@

# Reading in the data.
birth = read_csv(here::here("inputs/data/fig_1.csv"))

@@

birth1<-birth1 %>% 
  mutate(agegroup = as.character(str_replace(agegroup, "birthrate", "")))

@@

fig5_data <- fig5_data %>% 
  group_by(age_2024_year) %>% 
  mutate(cum_birthrate = cumsum(birthrate)/1000, 
         cohort = case_when(cohort2 == 1 ~ "1968-1972",
                                  cohort2 == 2 ~ "1973-1977",
                                  cohort2 == 3 ~ "1978-1982",
                                  cohort2 == 4 ~ "1983-1987",
                                  cohort2 == 5 ~ "1988-1992",
                                  cohort2 == 6 ~ "1993-1997"))

fig5_data <-  na.omit(fig5_data)
fig5_data %>%
  ggplot(aes(x = mage,
             y = cum_birthrate,
             color = cohort)) +
  geom_line()  +
  labs(
    title = "Children Ever Born by Mother’s Age and Birth Cohort",
    x = "Ages",
    y = "Cumulative birthrate",
    caption = "Source: NCHS Vital Statistics birth microdata and CDC SEER population data.",
    color = "Birth cohorts"
  ) +
  scale_x_continuous(limits=c(15, 44), breaks=seq(15, 44, by = 3)) +
  theme_minimal()+
  theme(legend.position = "right")

>>>>

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-2
#| fig-cap: Race and Ethnicity (ages 15-19 from 2000 to 2020)
birth1 = read_dta("data/nchs/nchs_births_pop_1990_2019.dta")

# getting total number of births for white people from 15 to 19 years old
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

# getting total number of births for black people from 15 to 19 years old  
birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

# getting total number of births for hispanics from 15 to 19 years old  
birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

# getting total number of births for white people from 20 to 24 years old
birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

# getting total number of births for black people from 20 to 24 years old
birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

# getting total number of births for hispanics from 20 to 24 years old  
birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

# gets the total number of births by year, compiled nationally
birthhist = birth1 %>%
  filter(year >= 2000) %>%
  group_by(year) %>% 
  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20 = sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20 = sum(numbirthhisp20))

# gets the birth rate for each race by adding the number of births from the 15 to 19for each race, then multiplying by 10,000 and dividing by the female population
birthhist %>%
  mutate(whiteall = white15*10000/sumpop, blackall= black15*10000/sumpop, hispall = hisp15*10000/sumpop) %>% 
  select(year, whiteall, blackall, hispall) %>%
  pivot_longer(cols = c("whiteall", "blackall", "hispall"),
               names_to = "racegroup",
               values_to = "value"
  ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000,2020, by=2))+
  labs(title = "birthrate for each race in age group 15 to 19",
       x = "Year",
       y = "Birth Rate",
       color = "Race") +
  scale_color_brewer(palette = "Set1", labels = c("Black (Non-Hispanic)", "Hispanic", "White (Non-Hispanic)")) +
  theme(legend.position = "bottom")

@@

birth1 = read_dta("data/nchs/nchs_births_pop_1990_2019.dta")

# getting total number of births for white people from 15 to 19 years old
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

# getting total number of births for black people from 15 to 19 years old  
birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

# getting total number of births for hispanics from 15 to 19 years old  
birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

# getting total number of births for white people from 20 to 24 years old
birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

# getting total number of births for black people from 20 to 24 years old
birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

# getting total number of births for hispanics from 20 to 24 years old  
birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

# gets the total number of births by year, compiled nationally
birthhist = birth1 %>%
  filter(year >= 2000) %>%
  group_by(year) %>% 
  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20 = sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20 = sum(numbirthhisp20))

# gets the birth rate for each race by adding the number of births from the 15 to 19for each race, then multiplying by 10,000 and dividing by the female population
birthhist %>%
  mutate(whiteall = white20*10000/sumpop, blackall= black20*10000/sumpop, hispall = hisp20*10000/sumpop) %>% 
  select(year, whiteall, blackall, hispall) %>%
  pivot_longer(cols = c("whiteall", "blackall", "hispall"),
               names_to = "racegroup",
               values_to = "value"
  ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000,2020, by=2))+
  labs(title="birthrate for each race in age group 20 to 24",
       x = "Year",
       y = "Birth Rate",
       color = "Race") +
  scale_color_brewer(palette = "Set1", labels = c("Black (Non-Hispanic)", "Hispanic", "White (Non-Hispanic)")) +
  theme(legend.position = "bottom")

@@

@@

# Reading in the data.

@@

birth1<-birth1 %>% 
  mutate(agegroup = as.character(str_replace(agegroup, "birthrate", "")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "1519", "15~19")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "2024", "20~24")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "2529", "25~29")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "3034", "30~34")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "3539", "35~39")))
birth1<-birth1 %>%  mutate(agegroup=as.character(str_replace(agegroup, "4044", "40~44")))

@@

fig5_data <- fig5_data %>% 
  group_by(age_2024_year) %>% 
  mutate(cum_birthrate = cumsum(birthrate)/1000, 
         cohort = case_when(cohort2 == 1 ~ "1968-1972        1992",
                                  cohort2 == 2 ~ "1973-1977        1997",
                                  cohort2 == 3 ~ "1978-1982        2002",
                                  cohort2 == 4 ~ "1983-1987        2007",
                                  cohort2 == 5 ~ "1988-1992        2012",
                                  cohort2 == 6 ~ "1993-1997        2017"))

fig5_data <-  na.omit(fig5_data)
fig5_data %>%
  ggplot(aes(x = mage,
             y = cum_birthrate,
             color = cohort)) +
  geom_line()  +
  labs(
    title = "Children Ever Born by Mother’s Age and Birth Cohort",
    x = "Ages",
    y = "Cumulative birthrate",
    caption = "Source: NCHS Vital Statistics birth microdata and CDC SEER population data.",
    color = "Birth cohorts   Age 20 in"
  ) +
  scale_x_continuous(limits=c(15, 44), breaks=seq(15, 44, by = 3)) +
  theme_minimal()+
  theme(legend.position = "right")
